---
title: "OA Overview"
author: "Thomas Klebel"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)

# setup -----
Sys.setenv(SPARK_HOME = "/usr/hdp/current/spark2-client")
library(sparklyr)
library(tidyverse)
library(arrow)
library(gganimate)
source(here::here("R/helpers.R"))

message("Connecting to spark...")

config <- spark_config()
config$spark.executor.cores <- 15
config$spark.executor.instances <- 5
config$spark.executor.memory <- "60G"
sc <- spark_connect(master = "yarn-client", config = config,
                    app_name = "SDG_OA_Markdown")
message("Connection to Spark successful!")

message("Reading the datasets...")
papers <- spark_read_csv(sc, "/user/tklebel/sdg/data/sdg_papers_collated.csv",
                         name = "papers")
affils <- spark_read_csv(sc,
                         "/user/tklebel/sdg/data/affiliations_with_country.csv",
                         name = "affils")
author_metadata <- spark_read_csv(sc,
                                  "/user/tklebel/sdg/data/sdg_author_data.csv",
                                  name = "author_metadata")
author_paper_affiliations <- spark_read_csv(
  sc,
  "/user/tklebel/sdg/data/sdg_author_paper_affil.csv",
  name = "author_paper_affiliations"
)
funded_projects <- spark_read_csv(
  sc,
  "/user/tklebel/sdg/data/openaire_funders_injoin_w_sdg.csv",
  name = "funded_projects"
)
wb_indicators <- spark_read_csv(
  sc,
  "/user/tklebel/sdg/data/world_bank_indicators.csv",
  name = "wb_indicators")
leiden <- spark_read_csv(sc,
                         "/user/tklebel/sdg/data/leiden_ranking.csv",
                         name = "leiden")

message("Successfully read all datasets!")
```


# OA by FOS and year
```{r, cache=TRUE}
# aggregate oa_status -----
oa_status <- papers %>%
  filter(!is.na(is_oa)) %>%
  select(paperid, fos_displayname, year, is_oa, oa_status) %>%
  distinct()


oa_per_year <- oa_status %>%
  count(fos_displayname, year, is_oa) %>%
  collect()

oa_status_per_year <- oa_status %>%
  count(fos_displayname, year, oa_status) %>%
  collect()
```



```{r sdg_oa_by_fos}
oa_per_year %>%
  group_by(fos_displayname, year) %>%
  mutate(oa_share = n/sum(n)) %>%
  filter(is_oa) %>%
  ggplot(aes(as.factor(year), oa_share, colour = fos_displayname,
             group = fos_displayname)) +
  geom_point() +
  geom_line() +
  scale_y_continuous(labels = function(x) scales::percent(x, accuracy = 1)) +
  labs(x = NULL, title = "OA share by Field of Study",
       y = NULL, colour = NULL) +
  theme(legend.position = "top")
```

```{r sdg_oa_type_by_fos}
oa_status_per_year %>%
  group_by(fos_displayname, year) %>%
  filter(!is.na(oa_status)) %>%
  mutate(oa_share = n/sum(n)) %>%
  ggplot(aes(lubridate::ymd(year, truncated = 2L), oa_share,
             colour = fos_displayname,
             group = fos_displayname)) +
  geom_point() +
  geom_line() +
  facet_wrap(vars(oa_status)) +
  scale_y_continuous(labels = function(x) scales::percent(x, accuracy = 1)) +
  labs(x = NULL, title = "OA share by Field of Study",
       y = NULL, colour = NULL) +
  theme(legend.position = "top")
```

The rise in OA is thus mainly due to rise in gold OA, slightly rise in hybrid.
However, unpaywall prefers gold over green: redo this figure with new 
categorisation.


# OA per funder 
```{r, cache=TRUE}
paper_oa_flag <- papers %>%
  select(paperid, is_oa, oa_status, year, fos_displayname)

oa_per_funder <- funded_projects %>%
  left_join(paper_oa_flag)

oa_per_funder_aggregated <- oa_per_funder %>%
  select(doi, funder_name, year, is_oa) %>%
  distinct() %>% # remove duplicate rows since many papers are funded by multiple projects
  group_by(year, funder_name) %>%
  count(is_oa) %>%
  filter(!is.na(is_oa)) %>%
  mutate(oa_share = n/sum(n),
         total_papers = sum(n)) %>%
  collect()
```

```{r sdg_oa_by_funder, fig.width=15, fig.height=12}
text_labels <- oa_per_funder_aggregated %>%
  group_by(funder_name) %>%
  summarise(nn_papers = sum(total_papers)) %>%
  mutate(label = glue::glue("n = {format(nn_papers, big.mark = '.', decimal.mark = ',')}"))


oa_per_funder_aggregated %>%
  filter(is_oa) %>%
  ggplot(aes(lubridate::ymd(year, truncated = 2L), oa_share,
             group = funder_name)) +
  geom_point() +
  geom_line() +
  geom_text(aes(x = lubridate::ymd("2017-01-01"),
                y = .4,
                label = label),
            data = text_labels) +
  facet_wrap(vars(str_wrap(funder_name, 50))) +
  scale_y_continuous(labels = function(x) scales::percent(x, accuracy = 1)) +
  labs(x = NULL, title = "OA share by funder",
       y = NULL, colour = NULL) +
  theme(legend.position = "top")
```

Share of OA going down for CA institutes of health. There is an explanation
for this in the Nature 2018 paper by Lariviere and Sugimoto.


# OA by country
```{r, cache=TRUE}
oa_per_affiliation <- paper_oa_flag %>%
  select(-fos_displayname) %>%
  left_join(author_paper_affiliations) %>%
  right_join(affils)

oa_per_affiliation_selected <- oa_per_affiliation %>%
  group_by(paperid) %>%
  mutate(frac_count = 1 / max(authorsequencenumber, na.rm = TRUE)) %>%
  select(paperid, authorid, is_oa, oa_status, year, country, frac_count) %>%
  filter(!is.na(is_oa))


oa_per_country <- oa_per_affiliation_selected %>%
  group_by(country, is_oa) %>%
  summarise(sum_frac_oa = sum(frac_count)) %>%
  mutate(prop_oa = sum_frac_oa/sum(sum_frac_oa)) %>%
  collect()


wb_local <- wb_indicators %>%
  filter(year == 2018) %>%
  collect()
```


## Correlation between indicators
```{r, fig.width=13, fig.height=13}
cor_matrix <- wb_local %>% 
  select(country_code, country_name, indicator_name, value) %>% 
  pivot_wider(names_from = "indicator_name", values_from = "value") %>% 
  select(-starts_with("country")) %>% 
  cor(use = "pairwise.complete.obs")

plot_correlation(cor_matrix, cluster = TRUE)
```



```{r oa_sdg_by_country_rd, fig.width=13, fig.height=9}
## OA per gdp, % spent on research
oa_with_r_d_gdp <- oa_per_country %>%
  left_join(wb_local, by = c("country" = "country_code")) %>%
  select(-indicator_name) %>%
  filter(indicator_code %in% c("GB.XPD.RSDV.GD.ZS", "NY.GDP.MKTP.KD")) %>%
  pivot_wider(names_from = indicator_code, values_from = value) %>%
  drop_na() %>%
  filter(is_oa)

oa_with_r_d_gdp %>%
  ggplot(aes(GB.XPD.RSDV.GD.ZS, prop_oa, colour = NY.GDP.MKTP.KD)) +
  geom_point(aes(size = NY.GDP.MKTP.KD), show.legend = FALSE) +
  scale_colour_viridis_c(trans = "log", option = "D", end = .9) +
  ggrepel::geom_text_repel(aes(label = country_name), show.legend = FALSE) +
  labs(x = "Expenditure towards research", y = "OA share") +
  scale_y_continuous(labels = scales::percent)
```


```{r oa_sdg_per_country_gdp_p_cap, fig.width=12, fig.height=8}
oa_with_gdp_per_cap <- oa_per_country %>%
  left_join(wb_local, by = c("country" = "country_code")) %>%
  select(-indicator_name) %>%
  filter(indicator_code %in% c("GB.XPD.RSDV.GD.ZS", "NY.GDP.PCAP.KD")) %>%
  pivot_wider(names_from = indicator_code, values_from = value) %>%
  drop_na() %>%
  filter(is_oa)


oa_with_gdp_per_cap %>%
  ggplot(aes(NY.GDP.PCAP.KD, prop_oa)) +
  geom_point() +
  ggrepel::geom_text_repel(aes(label = country_name)) +
  labs(x = "GDP per capita", y = "OA share") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_log10()
```


# OA by country with author groups
```{r, cache=TRUE}
author_paper_affiliations_w_groups <- author_paper_affiliations %>% 
  group_by(paperid) %>% 
  mutate(paper_author_cat = case_when(
    max(authorsequencenumber) == 1 ~ "single",
    max(authorsequencenumber) == 2 ~ "double",
    TRUE ~ "multi"
  )) %>% 
  mutate(author_position = case_when(
    paper_author_cat == "single" ~ "first_author",
    paper_author_cat == "double" & authorsequencenumber == 1 ~ "first_author",
    paper_author_cat == "double" & authorsequencenumber == 2 ~ "last_author",
    paper_author_cat == "multi" & authorsequencenumber == 1 ~ "first_author",
    paper_author_cat == "multi" & 
      authorsequencenumber == max(authorsequencenumber) ~ "last_author",
    TRUE ~ "middle_author"
  ))

oa_per_affiliation <- paper_oa_flag %>%
  select(-fos_displayname) %>%
  left_join(author_paper_affiliations_w_groups) %>%
  right_join(affils) %>% 
  filter(!is.na(is_oa))

oa_per_country_over_years <- oa_per_affiliation %>%
  count(country, year, author_position, is_oa) %>%
  group_by(country, year, author_position) %>% 
  mutate(prop_oa = n/sum(n)) %>%
  collect()

```


## author positions by country

```{r}
country_names <- wb_indicators %>% 
  distinct(country_name, country_code) %>% 
  collect()

pdata <- oa_per_country_over_years %>% 
  group_by(country, author_position) %>% 
  summarise(n = sum(n)) %>% 
  add_proportion(n, order_var = author_position,
                 order_string = "first") %>% 
  left_join(country_names, by = c("country" = "country_code")) %>% 
  drop_na()

p <- pdata %>% 
  ggplot(aes(fct_reorder(country_name, order), prop, fill = author_position)) +
  geom_col() +
  coord_flip() +
  labs(x = NULL, y = "Proportion")
plotly::ggplotly(p)
```

Here we could also look into the proportion of papers coming from single, dual 
or multi-author papers.


```{r sdg_oa_by_country_by_gdp, fig.width=8, fig.height=15}
gdp_p_cap <- wb_indicators %>% 
  filter(indicator_code == "NY.GDP.PCAP.KD") %>% 
  collect()

pdata <- oa_per_country_over_years %>% 
  group_by(country) %>% 
  mutate(country_with_low_n = any(n < 5)) %>% 
  filter(!country_with_low_n) %>% 
  left_join(gdp_p_cap, by = c("year", "country" = "country_code")) %>% 
  rename(gdp_p_cap = value) %>% 
  group_by(country, year, author_position) %>% 
  mutate(total_n = sum(n)) %>% 
  filter(is_oa)

pdata %>% 
  ggplot(aes(gdp_p_cap, prop_oa, size = total_n)) +
  geom_point() +
  facet_grid(rows = vars(year),
             cols = vars(author_position)) +
  # ggrepel::geom_text_repel(aes(label = country_name)) +
  labs(x = "GDP per capita", y = "OA share") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_log10()

```


### Only for first authors
```{r sdg_oa_by_country_by_gdp_first_authors, fig.height=10, fig.width=15}
pdata %>% 
  filter(author_position == "first_author") %>% 
  ggplot(aes(gdp_p_cap, prop_oa, size = total_n)) +
  geom_point() +
  facet_wrap(vars(year)) +
  # ggrepel::geom_text_repel(aes(label = country_name)) +
  labs(x = "GDP per capita", y = "OA share") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_log10()
```


## Animated and for all
```{r}
# use this answer to improve the approach: https://stackoverflow.com/a/54849281/3149349
p <- pdata %>%
  # filter(author_position == "first_author") %>%
  ggplot(aes(gdp_p_cap, prop_oa, size = n, group = country)) +
  geom_point() +
  facet_wrap(vars(author_position),
             nrow = 3) +
  # ggrepel::geom_text_repel(aes(label = country_name)) +
  labs(x = "GDP per capita", y = "OA share") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = 'Year: {frame_time}') +
  transition_time(year) +
  ease_aes('linear')
animate(p, renderer = ffmpeg_renderer())
```


```{r shut-down, echo=FALSE, message=FALSE}
message("Done. Shutting down...")
spark_disconnect(sc)
```

