---
title: "SDG OA author characteristics"
author: "Thomas Klebel"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# setup -----
Sys.setenv(SPARK_HOME = "/usr/hdp/current/spark2-client")
library(sparklyr)
library(tidyverse)
library(arrow)
source(here::here("R/helpers.R"))

message("Connecting to spark...")

config <- spark_config()
config$spark.executor.cores <- 15
config$spark.executor.instances <- 5
config$spark.executor.memory <- "60G"
sc <- spark_connect(master = "yarn-client", config = config,
                    app_name = "SDG_OA_authors")
message("Connection to Spark successful!")

message("Reading the datasets...")
papers <- spark_read_csv(sc, "/user/tklebel/sdg/data/sdg_papers_collated.csv",
                         name = "papers")
sdg_labels <- spark_read_csv(sc, "/user/tklebel/sdg/data/sdg_labels.csv",
                             name = "sdg_labels")
# join the labels here, since we are doing everything by SDG
papers <- papers %>% 
  left_join(sdg_labels)

affils <- spark_read_csv(sc,
                         "/user/tklebel/sdg/data/affiliations_with_country.csv",
                         name = "affils")
author_metadata <- spark_read_csv(sc,
                                  "/user/tklebel/sdg/data/sdg_author_data.csv",
                                  name = "author_metadata")
author_paper_affiliations <- spark_read_csv(
  sc,
  "/user/tklebel/sdg/data/sdg_author_paper_affil.csv",
  name = "author_paper_affiliations"
)
funded_projects <- spark_read_csv(
  sc,
  "/user/tklebel/sdg/data/openaire_funders_injoin_w_sdg.csv",
  name = "funded_projects"
)
wb_indicators <- spark_read_csv(
  sc,
  "/user/tklebel/sdg/data/world_bank_indicators.csv",
  name = "wb_indicators")
leiden <- spark_read_csv(sc,
                         "/user/tklebel/sdg/data/leiden_ranking.csv",
                         name = "leiden")

message("Successfully read all datasets!")
```


## Academic age
```{r, cache=TRUE, warning=FALSE}
author_w_year <- author_paper_affiliations %>%
  left_join(author_metadata) %>%
  select(paperid, authorid, authorsequencenumber, year_first_paper)

papers_w_age <- papers %>%
  # only take papers with OA status for now
  filter(!is.na(is_oa)) %>%
  left_join(author_w_year) %>%
  mutate(current_age = year - year_first_paper)


age_dist <- papers_w_age %>%
  filter(year == 2018) %>%
  distinct(authorid, current_age) %>%
  count(current_age) %>%
  collect()
```

```{r academic_age_overview}
age_dist %>%
  ggplot(aes(current_age, n)) +
  geom_col() +
  scale_y_continuous(labels = function(x) format(x,
                                                 big.mark = ".",
                                                 decimal.mark = ",")) +
  labs(y = "Number of researchers", title = "Current age of researchers in 2018",
       caption = "Only taking authors on papers that have an OA status into account")
```


```{r, cache=TRUE}
age_trunc <- papers_w_age %>%
  filter(current_age <= 50 | current_age > 0) %>%
  distinct(year, paperid, is_oa, oa_status, current_age, authorsequencenumber)

# make age groups
age_oa <- age_trunc %>%
  mutate(age_group = case_when(
    current_age > 0 & current_age < 10 ~ "1-9 years",
    current_age >= 10 & current_age < 20 ~ "10-19 years",
    current_age >= 20 & current_age < 30 ~ "20-39 years",
    current_age >= 30 & current_age < 40 ~ "30-39 years",
    current_age >= 40 & current_age < 50 ~ "40-49 years",
    current_age >= 50 & current_age < 60 ~ "50-59 years",
    TRUE ~ "60+ years"
  )) %>%
  group_by(year, age_group) %>%
  summarise(oa_share = mean(as.numeric(is_oa))) %>%
  collect()
```


```{r oa_authors_by_academic_age, fig.height=8, fig.width=10}
age_oa %>%
  ggplot(aes(as_year(year), oa_share, colour = age_group, group = age_group)) +
  geom_point() +
  geom_line() +
  scale_y_continuous(labels = function(x) scales::percent(x, accuracy = 1)) +
  theme_bw() +
  labs(x = NULL, y = "OA share", title = "OA share over time", colour = NULL)
```


# Group into first, last, and middle author
```{r, cache=TRUE}
author_paper_affiliations_w_groups <- make_author_groups(author_paper_affiliations)

age_base <- papers %>% 
  select(paperid, SDG_label, year, is_oa) %>% 
  left_join(author_paper_affiliations_w_groups) %>% 
  left_join(author_metadata) %>% 
  select(paperid, SDG_label, year, author_position, authorid, 
         year_first_paper, is_oa, paper_author_cat)

age_cohorts <- age_base %>% 
  group_by(year, authorid) %>% 
  mutate(current_age = year - year_first_paper) %>% 
  filter(current_age > 0) %>% 
  mutate(age_cohort = case_when(
    year_first_paper < 1960 ~ NA_character_,
    year_first_paper < 1970 ~ "1960-1969",
    year_first_paper < 1980 ~ "1970-1979",
    year_first_paper < 1990 ~ "1980-1989",
    year_first_paper < 2000 ~ "1990-1999",
    year_first_paper < 2010 ~ "2000-2009",
    TRUE ~ "2010-2017"
  )) %>% 
  ungroup()

age_oa_cohort <- age_cohorts %>% 
  group_by(year, age_cohort, SDG_label) %>%
  summarise(oa_share = mean(as.numeric(is_oa))) %>%
  collect()
```

```{r oa_authors_by_academic_age_by_SDG, fig.height=7, fig.width=10}
age_oa_cohort %>%
  filter(age_cohort != "NA") %>% 
  ggplot(aes(as_year(year), oa_share, colour = SDG_label)) +
  geom_point() +
  geom_line() +
  facet_wrap(vars(age_cohort)) +
  scale_y_continuous(labels = function(x) scales::percent(x, accuracy = 1)) +
  theme_bw() +
  labs(x = NULL, y = "OA share", title = "OA share by year of first paper",
       colour = NULL)
```

```{r oa_authors_by_academic_age_by_SDG, fig.height=10, fig.width=10}
age_oa_cohort %>%
  filter(age_cohort != "NA") %>% 
  ggplot(aes(as_year(year), oa_share, colour = age_cohort)) +
  geom_point() +
  geom_line() +
  facet_wrap(vars(SDG_label), nrow = 2) +
  scale_y_continuous(labels = function(x) scales::percent(x, accuracy = 1)) +
  theme_bw() +
  theme(legend.position = c(.8, .2)) +
  labs(x = NULL, y = "OA share", title = "OA share over time", colour = NULL)
```

## By author position
```{r}
age_oa_positions <- age_cohorts %>% 
  # only consider multi author papers
  filter(paper_author_cat == "multi") %>% 
  group_by(year, age_cohort, SDG_label, author_position) %>%
  summarise(oa_share = mean(as.numeric(is_oa)),
            n = n()) %>%
  collect()
```

```{r oa_authors_by_academic_age_by_SDG_by_position, fig.height=10, fig.width=10}
age_oa_positions %>%
  filter(age_cohort != "NA", year < 2020) %>% 
  ggplot(aes(as_year(year), oa_share, colour = age_cohort)) +
  geom_point() +
  geom_line() +
  facet_grid(rows = vars(SDG_label), cols = vars(author_position)) +
  scale_y_continuous(labels = function(x) scales::percent(x, accuracy = 1)) +
  theme_bw() +
  theme(legend.position = c(.9, .1)) +
  labs(x = NULL, y = "OA share", title = "OA share over time", colour = NULL,
       caption = "Only considering multi author papers. Dropping 2020 because of low cell counts")
```

Findings:

- MEdicine for first authors: very narrow bandwith, compared to last and middle
authors - age is not very relevant there. 
- medicine last authors: younger researchers (from 2000) initially considerably
lower OA rate, more or less caught up with the rest
- Overall: OA share is lower for younger authors

# Individual prestige
Issue: the current perspective is agnostic to the timeline: citations are 
aggregated towards each author. To make a claim about whether being prestigious
makes you publish OA we would need to know the citations prior to publication.
but this data is not available in MAG. It could be generated from the raw graph:
counting the links to each paper per year. But probably out of scope for now.
```{r}

```

## Citations
## Publications


# Institutional prestige
## Citations
## Publications




```{r shut-down, echo=FALSE, message=FALSE}
message("Done. Shutting down...")
spark_disconnect(sc)
```
