---
title: "05-gender-overview"
author: "Thomas Klebel"
date: "8 7 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(sparklyr)
library(tidyverse)
library(arrow)

source(here::here("R/helpers.R"))

theme_set(theme_bw())

unique_names <- read_csv(here::here("data/processed/unique_names.csv"))
genderized_names <- read_csv(here::here("data/processed/genderized_names.csv"))



message("Connecting to spark...")

Sys.setenv(SPARK_HOME = "/usr/hdp/current/spark2-client")
config <- spark_config()
config$spark.executor.cores <- 15
config$spark.executor.instances <- 3
config$spark.executor.memory <- "60G"
sc <- spark_connect(master = "yarn-client", config = config,
                    app_name = "SDG_OA_authors")
message("Connection to Spark successful!")

message("Reading the datasets...")
author_metadata <- spark_read_csv(sc,
                                  "/user/tklebel/sdg/data/sdg_author_data.csv",
                                  name = "author_metadata")

```


# How many names were genderized?
```{r}
uniqe_names %>% 
  left_join(genderized_names, by = c("first_name" = "name")) %>% 
  summarise(n = n(),
            n_genderized = sum(!is.na(gender)),
            prop_genderized = n_genderized/n)
```


```{r, cache=TRUE}
sampling_grid <- expand_grid(counts = 1:100,
                             prob = seq(.5, 1, by = .1))


get_n <- function(df, counts, prob) {
  df %>% 
    filter(count >= counts, probability >= prob) %>% 
    nrow()
}


res <- sampling_grid %>% 
  mutate(resulting_n = map2_int(counts, prob, ~get_n(genderized_names, .x, .y)))

res %>% 
  head(20) %>% 
  knitr::kable()

```

```{r}
pdata <- res %>% 
  mutate(prop = resulting_n/nrow(genderized_names))

pdata %>% 
  ggplot(aes(counts, prop, colour = as.factor(prob))) +
  geom_point() +
  scale_y_continuous(labels = scales::percent)
p
```


```{r}
p +
  coord_cartesian(xlim = c(0, 20))
```

In terms of unique names, setting count higher than abou 3-5 already leads to 
way less names being covered. From this I would maybe choose to include names
that appear more than three times and have at least 80% probability. This would
result in only including about 50% of the genderized names.

However: how does this translate to actual authors in our dataset?
